Left-linear rules in adhesive categories

### 2024-02-13 icalp/fusione.tex
cose
minimissima

\begin{abstract}
%  When two sequences of steps executed by a computational device can
%  be considered equivalent?  This is a relevant question if one
%  considers concurrent and distributed systems, since the independence
%  of two steps means that they could be executed in any order, and
%  potentially in parallel.
  When can two sequential steps performed by a computing device be
  considered (causally) independent?  This is a relevant question for
  concurrent and distributed systems, since independence means that
  they could be executed in any order, and potentially in parallel.
  % 
  We investigate the issue in the context of adhesive categories, a
  very general instance of the DPO approach to rewriting. More
  precisely, we focus on left-linear rules, which not only consume,
  preserve and generate entities, as linear rules do, but may
  ``merge'' parts of the state.
  % 
  An apparently minimal, yet technically hard enhancement, 
  % 
  % We refine the established notion of sequential independence
  % between steps, and ...
  since the canonical notion of independence used for linear rewriting
  systems does not work out of the box, but it requires the
  restriction to well-behaved classes of left-linear rules.
  %
  Moreover, a standard characterisation of independence that -- in the
  linear case -- allows one to derive a number of properties,
  essential in the development of a theory of concurrency, no longer
  holds.
  %and this requires a substantial revision of the theory.
  %
  The paper performs an in-depth study of the notion of independence
  for left-linear rules: it introduces a novel characterisation of
  independence, identifies well-behaved classes of left-linear
  rewriting systems and provides some fundamental results including a
  Church-Rosser property and the existence of canonical equivalence
  proofs for concurrent computations. These results properly extends
  the class of formalisms that can be modelled in the adhesive
  framework.
\end{abstract}

@@

Among rewriting-based formalisms, those manipulating graph-like structures
proved useful in many settings for representing the dynamics of distributed systems: the 
graph serves to represent the entities and their relations within states, while 
rewriting steps, modifying the graph, model computation steps that result 
in changes to the system state. The DPO (double-pushout) approach~\cite{CMREHL:AAGT} 
is nowadays considered the standard one for these structures, thanks to its locality 
(rule application has a local effect on a state, differently e.g. from the SqPO 
approach~\cite{CorradiniHHK06}) and most importantly to its flexibility: the rules allow to 
specify that,  in a rewriting step, some entities in the current state are removed, 
some others are needed for the rewriting step to occur but remain unaltered and 
some new entities can be generated. Concerning concurrency, the chosen notion 
is shift equivalence, and independence is formally captured by what is called the
Church-Rosser theorem of DPO rewriting~\cite[Section~xxx]{CMREHL:AAGT}.

@@

So far, most of the theoretical results focussed on what are called
\emph{linear rules}: intuitively, a step is required only to consume,
preserve and generate entities.  However, in some situations it is
also necessary to ``merge'' parts of the state.  This happens already
in the context of string diagram mentioned above, but also in
graphical implementations of nominal calculi where, as a result of
name passing, the received name is identified with a local one at the
receiver~\cite{CVY:ESSPE,Gad07} or in the visual modelling of bonding
in biological/chemical processes~\cite{PUY:MBPE}. A similar mechanisms
is at the core of e-graphs (equality graphs), recently used for
rewrite-driven compiler optimisations~\cite{WNW:egg}: rather than
explicitly merging nodes corresponding to equal expression, the idea
here -- conceptually and, as we will see, also mathematically similar
-- is to maintain an
equivalence %\todo{P: more precisely a congruence}
over the nodes.

Technically, to acquire the expressiveness needed for capturing
fusions, i.e. the merge of some elements in the state, requires to
move from linear to left-linear rewriting rules. While left-linear
rules have been considered in various instances (mainly in categories
of graphs) and some results concerning the corresponding rewriting
theory have been put forward, to the best of our knowledge the
phenomena arising when devising a theory of independence and
concurrency for rewriting with left-linear rules in the general
setting of adhesive and quasi-adhesive categories
%(which capture most ``applications'') 
have not been systematically
explored~\cite{Ehrig1976,EHP:BRfTToHLRS,baldan2011adhesivity},
even if an investigation in the setting of quasi-topoi, considering general
non-linear rules yet restricting their applicability,
% of rules and focussing as far as DPO in concerned on quasi-adhesivity, 
has been presented in~\cite{BehrHK23}.

@@

The following proposition establishes a key property of $\mathcal{A}$-Van Kampen squares with a mono as a side: they are not only pushouts, but also pullbacks \cite{ehrig2004adhesive,behr2022fundamentals,lack2005adhesive}.

>>>>

\begin{abstract}
%  When two sequences of steps executed by a computational device can
%  be considered equivalent?  This is a relevant question if one
%  considers concurrent and distributed systems, since the independence
%  of two steps means that they could be executed in any order, and
%  potentially in parallel.
  When can two sequential steps performed by a computing device be
  considered (causally) independent?  This is a relevant question for
  concurrent and distributed systems, since independence means that
  they could be executed in any order, and potentially in parallel.
  % 
  We investigate the issue in the context of adhesive categories, a
  very general instance of the DPO approach to rewriting.\todo{[A] direi: ``...
    in the context of the Double Pushout (DPO) Approach to rewriting in adhesive categories.''}
    More
  precisely, we focus on left-linear rules, which not only consume,
  preserve and generate entities, as linear rules do, but may
  ``merge'' parts of the state.
  % 
  An apparently minimal, yet technically hard enhancement, 
  % 
  % We refine the established notion of sequential independence
  % between steps, and ...
  since the canonical notion of independence used for linear rewriting
  systems does not work out of the box, but it requires the
  restriction to well-behaved classes of left-linear rules.
  %
  Moreover, a standard characterisation of independence that -- in the
  linear case -- allows one to derive a number of properties,
  essential in the development of a theory of concurrency, no longer
  holds.
  %and this requires a substantial revision of the theory.
  %
  The paper performs an in-depth study of the notion of independence
  for left-linear rules: it introduces a novel characterisation of
  independence, identifies well-behaved classes of left-linear
  rewriting systems and provides some fundamental results including a
  Church-Rosser property and the existence of canonical equivalence
  proofs for concurrent computations. These results properly extends
  the class of formalisms that can be modelled in the adhesive
  framework.
\end{abstract}

@@

Among rewriting-based formalisms, those manipulating graph-like structures
proved useful in many settings for representing the dynamics of distributed systems: the 
graph serves to represent the entities and their relations within states, while 
rewriting steps, modifying the graph, model computation steps that result 
in changes to the system state. The DPO (double-pushout) approach~\cite{EhrigPS73}
is nowadays considered a standard one for these structures, thanks to its locality 
(rule application has a local effect on a state, differently from more recent approaches
like, e.g., SqPO~\cite{CorradiniHHK06}
or PBPO$^+$~\cite{OverbeekER23}) and to its flexibility: the rules allow to 
specify that,  in a rewriting step, some entities in the current state are removed, 
some others are needed for the rewriting step to occur but remain unaltered and 
some new entities can be generated. Concerning concurrency, the chosen notion 
is shift equivalence, and independence is formally captured by what is called the
Church-Rosser theorem of DPO rewriting~\cite[Section~4]{Ehrig78}.

@@

So far, most of the theoretical results focussed on what are called
\emph{linear rules}: intuitively, a step is required only to consume,
preserve and generate entities.  However, in some situations it is
also necessary to ``merge'' parts of the state.  This happens already
in the context of string diagrams mentioned above, but also in
graphical implementations of nominal calculi where, as a result of
name passing, the received name is identified with a local one at the
receiver~\cite{CVY:ESSPE,Gad07}, or in the visual modelling of bonding
in biological/chemical processes~\cite{PUY:MBPE}. A similar mechanisms
is at the core of e-graphs (equality graphs), recently used for
rewrite-driven compiler optimisations~\cite{WNW:egg}: rather than
explicitly merging nodes corresponding to equal expressions, the idea
here -- conceptually and, as we will see, also mathematically similar
-- is to maintain an
equivalence %\todo{P: more precisely a congruence}
over the nodes.

Technically, to acquire the expressiveness needed for capturing
fusions, i.e. the merge of some elements in the state, one is required to
move from linear to left-linear rewriting rules. While left-linear
rules have been considered in various instances (mainly in categories
of graphs) and some results concerning the corresponding rewriting
theory have been put forward, to the best of our knowledge the
phenomena arising when devising a theory of independence and
concurrency for rewriting with left-linear rules in the general
setting of adhesive and quasi-adhesive categories
%(which capture most ``applications'') 
have not been systematically
explored~\cite{Ehrig1976,EHP:BRfTToHLRS,baldan2011adhesivity},
even if an investigation in the setting of quasi-topoi, considering general
non-linear rules yet restricting their applicability,
% of rules and focussing as far as DPO in concerned on quasi-adhesivity, 
has been presented in~\cite{BehrHK23}.

@@

The following proposition establishes a key property of $\mathcal{A}$-Van Kampen squares with a mono as a side: they are not only pushouts, but also pullbacks \cite{ehrig2004adhesive,BehrHK23,lack2005adhesive}.
